\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-17 papers/jfp/1-intro.tex
Minor fixes and text tweaks, mostly for nicer formatting

\item We can use these new abstractions to more clearly explain details from the
original \Shake paper~\cite{mitchell2012shake} (\S\ref{sec-step-traces},
\S\ref{sec-experience-shake}) and build new features useful for the GHC build
system~\cite{hadrian} (\S\ref{sec-using-cloud-shake}).

@@

In short, instead of seeing build systems as unrelated points in space, we now
see them as locations in a connected landscape, leading to a better
understanding of what they do and how they compare, and making it easier to
explore other points in the landscape. While we steer clear of many engineering
aspects of real build systems, in~\S\ref{sec-engineering} we show how they can
be incorporated into the presented abstractions. The related work is discussed
in~\S\ref{sec-related}.

This paper is an extended version of an earlier conference
paper~\cite{mokhov2018buildsystems}. The key changes compared to the earlier
version are: (i) we added further clarifications and examples
into~\S\ref{sec-abstractions}, in particular,~\S\ref{sec-task-examples} is
entirely new; (ii)~\S\ref{sec-scheduler} and~\S\ref{sec-rebuilder} are based on
the material from the conference paper but have been substantially expanded to
include further details and examples, as well as completely new material such
as~\S\ref{sec-step-traces}; (iii)~\S\ref{sec-experience} is completely new; (iv)
\S\ref{sec-failures} and \S\S\ref{sec-polymorphism}-\ref{sec-tracking-aspects}
are almost entirely new, and \S\ref{sec-non-determinism} has been revised.
The new material focuses on our experience and various important practical
considerations, hence justifying the ``and Practice'' part of the paper title.

>>>>

\item We can use the presented abstractions to more clearly explain details from
the original \Shake paper~(\S\ref{sec-step-traces}, \S\ref{sec-experience-shake})
and develop new cloud build features, which are already in use in industry and in
the GHC build system~(\S\S\ref{sec-using-cloud-shake}-\ref{sec-hadrian}).

@@

In short, instead of seeing build systems as unrelated points in space, we now
see them as locations in a connected landscape, leading to a better
understanding of what they do and how they compare, and making it easier to
explore other points in the landscape. While we steer clear of many engineering
aspects of real build systems, in~\S\ref{sec-engineering} we discuss these
aspects in the context of the presented abstractions. The related work is covered
in~\S\ref{sec-related}.

This paper is an extended version of an earlier conference
paper~\cite{mokhov2018buildsystems}. The key changes compared to the earlier
version are: (i) we added further clarifications and examples
to~\S\ref{sec-abstractions}, in particular,~\S\ref{sec-task-examples} is
entirely new; (ii)~\S\ref{sec-scheduler} and~\S\ref{sec-rebuilder} are based on
the material from the conference paper but have been substantially expanded to
include further details and examples, as well as completely new material such
as~\S\ref{sec-step-traces}; (iii)~\S\ref{sec-experience} is completely new; (iv)
\S\ref{sec-failures} and \S\S\ref{sec-polymorphism}-\ref{sec-file-watching}
are almost entirely new, and \S\ref{sec-non-determinism} has been revised.
The new material focuses on our experience and various important practical
considerations, hence justifying the ``and Practice'' part of the paper title.

### 2019-10-17 papers/jfp/10-conclusions.tex
Minor fixes and text tweaks, mostly for nicer formatting

Andrey Mokhov is funded by a Royal Society Industry Fellowship \cmd{IF160117}
on the topic ``Towards Cloud Build Systems with Dynamic Dependency Graphs''.

>>>>

Andrey Mokhov's research was funded by a Royal Society Industry Fellowship
\cmd{IF160117} on the topic ``Towards Cloud Build Systems with Dynamic
Dependency Graphs''.

### 2019-10-17 papers/jfp/2-background.tex
Minor fixes and text tweaks, mostly for nicer formatting

Build systems automate the execution of repeatable tasks, at a scale from individual users
up to large organisations. In this section we explore the design space of build
systems, using four concrete examples: \Make~\cite{feldman1979make},
\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
\Excel~\cite{advanced_excel}.
We have carefully chosen these four to illustrate the various axes on which
build systems differ; we discuss many other notable examples of build systems,
and their relationships, in~\S\ref{sec-implementations} and~\S\ref{sec-related}.

@@

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification times} to detect which files changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}. For a more concrete description
see~\S\ref{sec-implementation-make}.

@@

To support dynamic dependencies, \Excel's calculation engine~\cite{excel_recalc}
is significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value or formula is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

The dependencies of \cmd{release.tar} are not known statically: they are
determined by the content of \cmd{release.txt}, which might not even exist
before the build. Makefiles cannot express such dependencies, requiring problematic
workarounds such as \emph{build phases}~\cite{hadrian}.
In \Shake we can express the rule for \cmd{release.tar} as:

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt,
it uses the dependency graph that is dynamically constructed during
the previous build. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not (see \S\ref{sec-verifying-traces} for an explanation of why).

@@

\vspace{-1.5mm}
\subsection{Summary}\label{sec-background-summary}
\vspace{-0.5mm}

>>>>

Build systems automate the execution of repeatable tasks, at a scale from individual
users up to large organisations. In this section we explore the design space of build
systems, using four examples: \Make~\cite{feldman1979make},
\Shake~\cite{mitchell2012shake}, \mbox{\Bazel~\cite{bazel}}, and
\Excel~\cite{advanced_excel}.
We have carefully chosen these four to illustrate the various axes on which
build systems differ; we discuss many other notable examples of build systems,
and their relationships, in~\S\ref{sec-implementations} and~\S\ref{sec-related}.

@@

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification times} to detect which files changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}. For a more concrete description
see~\S\ref{sec-topological} and~\S\ref{sec-implementation-make}.

@@

To support dynamic dependencies, \Excel's calculation engine~\cite{excel_recalc}
is significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value or formula is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and discuss it
in more detail in~\S\ref{sec-restarting} and~\S\ref{sec-implementation-excel}.

@@

The dependencies of \cmd{release.tar} are not known statically: they are
determined by the content of \cmd{release.txt}, which might not even exist
before the build. Makefiles cannot express such dependencies, requiring
workarounds such as \emph{build phases}, which are known to be
problematic~\cite{hadrian}. In \Shake we can express the rule for
\cmd{release.tar} as:

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt,
it uses the dependency graph that is dynamically constructed during
the previous build. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, as well as
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a further discussion
in~\S\ref{sec-suspending} and a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not; see~\S\ref{sec-dirty-bit} for an explanation of why.

@@

% \vspace{-1.5mm}
\subsection{Summary}\label{sec-background-summary}
% \vspace{-0.5mm}

### 2019-10-17 papers/jfp/3-abstractions.tex
Add 7.1 on Haskell as design language
Note that newer Docker builds aren't merely Functor anymore
Minor fixes and text tweaks, mostly for nicer formatting

\vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
\vspace{-0.5mm}

@@

\begin{figure}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler (see §4) and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
% \vspace{-3.5mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
% \vspace{-4.5mm}
\end{figure}

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task  c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}

@@

\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
\end{minted}

@@

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

>>>>

% \vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
% \vspace{-0.5mm}

@@

\begin{figure}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
run :: c f => Task c k v -> (k -> f v) -> f v
run (Task task) fetch = task fetch
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler (see §4) and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
% \vspace{-3.5mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
% \vspace{-4.5mm}
\end{figure}

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}

@@

\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}

@@

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}
  (ignoring the recent multi-stage builds). It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

### 2019-10-17 papers/jfp/5-rebuilders.tex
Minor fixes and text tweaks, mostly for nicer formatting

With a dirty bit it is possible to achieve
minimality~(\S\ref{sec-background-make}). However, to achieve early
cutoff~(\S\ref{sec-background-shake}) it would be important to clear the dirty
bit after a computation that did not change the value and make sure that cells
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
keys that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.

A dirty bit rebuilder is useful to reduce memory consumption, and in the case
of \Make, to integrate with the file system. However, as the examples show, in
constrained environments where a dirty bit is chosen, it is often done as part
of a series of compromises.

It is possible to write a dirty-bit builder which is minimal and supports early
cutoff. To do so, you need to start with all inputs that have changed marked dirty,
then a node must be rebuilt if any children are dirty, and must set that node to dirty
only if the result changes. At the end of the build all dirty bits
must be reset to not-dirty. This approach only works if all targets are rebuilt each time,
since clearing dirty bits of children who aren't targets will cause them to incorrectly
not rebuild subsequently. To avoid resetting the dirty bit,
it is possible to use successive execution numbers, which ultimately leads to an approach
we call verifying step traces in \S\ref{sec-step-traces}.

@@

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild~\S\ref{sec-implementation-cloud}, can index directly from the
key and the hashes of its dependencies to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies locally, and then make a single call to the server to retrieve
the result.

>>>>

With a dirty bit it is possible to achieve
minimality~(\S\ref{sec-background-make}). However, to achieve early
cutoff~(\S\ref{sec-background-shake}) it would be important to clear the dirty
bit after a computation that did not change the value and make sure that keys
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
files that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.

A dirty-bit rebuilder is useful to reduce memory consumption, and in the case
of \Make, to integrate with the file system. However, as the examples show, in
constrained environments where a dirty bit is chosen, it is often done as part
of a series of compromises.

It is possible to implement a dirty-bit rebuilder that is minimal and supports early
cutoff. To do so, the build system should start with all inputs that have changed
marked dirty, then a key must be rebuilt if any of its direct dependencies are dirty,
marking the key dirty only if the result has changed. At the end of the build all
dirty bits must be cleared. This approach only works if all targets are rebuilt each
time because clearing dirty bits of keys that are not transitive dependencies of
current targets will cause them to incorrectly not rebuild subsequently. To avoid
resetting the dirty bit, it is possible to use successive execution numbers, which
ultimately leads to an approach we call verifying step traces in~\S\ref{sec-step-traces}.

@@

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild~(\S\ref{sec-implementation-cloud}), can index directly from the
key and the hashes of its dependencies to the output result, i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies locally, and then make a single call to the server to retrieve
the result.

### 2019-10-17 papers/jfp/6-implementations.tex
Add 7.1 on Haskell as design language
Minor fixes and text tweaks, mostly for nicer formatting

\subsection{Concrete Implementations}

@@

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs
with constructive traces~\cite{esfahani2016cloudbuild}. Note that
we need to convert a monadic \hs{ctRebuilder} into an applicative one by
applying an adapter function \hs{adaptRebuilder}, which unwraps a given
\hs{Task}~\hs{Applicative} and wraps it into \hs{Task}~\hs{Monad}.

>>>>

\subsection{Concrete Implementations} \label{sec-concrete-impl}

@@

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs
with constructive traces~\cite{esfahani2016cloudbuild}. We convert a
monadic \hs{ctRebuilder} into an applicative one by applying an adapter
\hs{adaptRebuilder}, which unwraps a given \hs{Task}~\hs{Applicative}
and wraps it into \hs{Task}~\hs{Monad}.

### 2019-10-17 papers/jfp/7-experience.tex
Typo
Minor fixes and text tweaks, mostly for nicer formatting

Moreover, doing so is extremely beneficial.
The discipline of writing executable prototypes in
Haskell had a profound effect on our thinking. It forced misconceptions
to to surface early.  It required us to be explicit about side effects.
It gave us a huge incentive to design abstractions that had simple types
and an explicable purpose.

Consider, for example, our \hs{Task} type:
\begin{minted}[xleftmargin=10pt]{haskell}
data Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}
We started off with a much more concrete type, and explored
multiple variants.  During those iterations, this single line of
code gave us a tangible, concrete basis for productive conversations,
much more so than general conversations about ``tasks''.

It is also worth noting that we needed a rather expressive language,
to faithfully express the abstractions that seen natural in this
setting.  In the case of \hs{Task} we use a data constructor with a
polymorphic field; a higher-kinded type variable \hs{f :: * -> *}; an
even more abstracted kind for \hs{c :: (* -> *) -> Constraint}; and,
of course, type classes.

@@

\subsection{Experience from building GHC with \Shake}

>>>>

Moreover, doing so is extremely beneficial.
The discipline of writing executable prototypes in
Haskell had a profound effect on our thinking. It forced misconceptions
to surface early.  It required us to be explicit about side effects.
It gave us a huge incentive to design abstractions that had simple types
and an explicable purpose.

Consider, for example, our \hs{Task} type:
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}
We started off with a much more concrete type, and explored
multiple variants.  During those iterations, this single line of
code gave us a tangible, concrete basis for productive conversations,
much more so than general conversations about ``tasks''.

It is also worth noting that we needed a rather expressive language,
to faithfully express the abstractions that seem natural in this
setting.  In the case of \hs{Task} we use a data constructor with a
polymorphic field; a higher-kinded type variable \hs{f}~\hs{::}~\hs{*}~\hs{->}~\hs{*}; an
even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
of course, type classes.

@@

\subsection{Experience from building GHC with \Shake}\label{sec-hadrian}

### 2019-10-17 papers/jfp/8-engineering.tex
Minor fixes and text tweaks, mostly for nicer formatting

Overall we choose not to deal with cyclic dependencies -- a choice that most build
systems also follow. There are computation frameworks that support tasks with
cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
certain sense~\cite{pottier2009lazy,radul2009propagation}.

@@

newtype TaskT c k v = TaskT@\,@{@\,@runT :: @\std{forall}@ f. c f => Fetch k f -> f v@\,@}

@@

\subsection{File watching vs polling}\label{sec-integration-with-ides}

>>>>

We choose not to deal with cyclic dependencies -- a choice that most build
systems also follow. There are computation frameworks that support dependency
cycles under the assumption that tasks are \emph{monotonic} in a certain
sense~\cite{pottier2009lazy,radul2009propagation}.

@@

newtype TaskT c k v = TaskT (@\std{forall}@ f. c f => Fetch k f -> f v)

@@

\subsection{File watching vs polling}\label{sec-file-watching}
